---
title: 'Threshold of Risk Inference: Examples with Simulated Data'
author: "Kevin Donovan"
date: June 13, 2019
output: html_document
---

The repository found at

https://github.com/kmdono02/Risk_Threshold

includes source code for R functions which implement the nonparametric methods for inference on risk thresholds detailed 
in "Nonparametric inference for immune response thresholds of risk in vaccine studies" by Donovan, Hudgens, and Gilbert (2018).  Below, some examples of these methods are illustrated using simulated data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_packages, include=FALSE}
library(tidyverse)
library(plotrix) # used to plot confidence intervals and estimates using function plotCI
```

# Cross-sectional Data
## Positive risk levels
Let's consider the scenario where inference on risk thresholds is done with cross-sectional data collected at a single time point (as detailed in Sections 2 and 3 of the paper).  Let $Y$ denote disease status ($Y=1$ denoting disease, $Y=1$ denoting no disease) and $S$ denote immune response.  Suppose $S$ is observed for proportion $p_0$ of subjects with $Y=0$ and proportion $p_1$ of subjects with $Y=1$.  In the following simulated examples, sample size $n=12,500$ where $S$ is generated from a gamma distribution with mean and variance 4 and $Y|S$ is generated from a Bernoulli distribution with $\Pr(Y=1|S)=$exbit$(\beta_0+\beta_1S)$.  The slope $\beta_1$ was set to -5 and $\beta_0$ was determined based on the chosen disease risk $\Pr(Y=1)$; in the following simulations risks of 0.01 and 0.1 were used with $p_0=0.2$ and $p_1=1$.  The R function sim_data_logistic found in the GitHub repository simulates data under this model (based on choosen values of $n, \beta_1, t_0$, $p_0$, $p_1$, and disease risk).  

For disease risk 0.01, point estimates and 95$\%$ confidence intervals are computed of thresholds for risk levels 0.001, 0.003, 0.005, 0.007, and 0.009 and for disease risk 0.1, point estimates and 95$\%$ confidence intervals are computed of thresholds for risk levels 0.009, 0.01, 0.03, 0.05, 0.07, and 0.09.  The confidence intervals are computed using 2000 bootstrap resamples.  These calculations were done using the R function Thres_Surv, whose source code can be found in the GitHub repository, which implements the methods detailed for Section 4 for inference with time to event data.  These estimates and confidence intervals were computed using $\hat{v}_c^{w}$ as defined in Section 3 of the paper.  For each threshold, the point estimate and confidence interval using $\hat{v}_c^{w}$ are plotted with the corresponding risk level on x axis; the point estimate is colored red and the true threshold is colored blue.

The true thresholds under the model from which the data were simulated (models detailed previously) are also provided (function sim_data_logit calculates these for vector of choosen risk levels).

```{r cross_sec_logit}
source("../sim_data_logit.R")
source("../Thres_CrossSect.R")

## Low risk example
sim_data <- Sim_data_logit(Risk = 0.01, n=12500, p=c(0.001,0.003,0.005,0.007,0.009), p0=0.2, p1=1, seed_val = 012)

vac_data_analysis_LR <- sim_data$simmed_data %>% select(S_obs, Y) 
low_risk_cross_sec_results <- Thres_CrossSect(dataset = vac_data_analysis_LR, p=c(0.001,0.003,0.005,0.007,0.009), weighted=TRUE, bts=2000)
low_risk_cross_sec_results

# Now plot
plotCI(x=low_risk_cross_sec_results$Risk_Level, y=low_risk_cross_sec_results$Est, 
       ui=low_risk_cross_sec_results$CI_UL,li=low_risk_cross_sec_results$CI_LL, xlab="Risk Level", ylab="Threshold Estimate", col="red",scol="black", main="Threshold of risk inference on simulated data: disease risk=0.01\nCross sectional scenario", ylim=c(0.25, 1.75))
points(x=as.numeric(names(sim_data$true_thresholds)), y=sim_data$true_thresholds, col="blue")

## Medium risk example
sim_data <- Sim_data_logit(Risk = 0.1, n=12500, p=c(0.009, 0.01, 0.03, 0.05, 0.07, 0.09), p0=0.2, p1=1, seed_val = 012)

vac_data_analysis_MR <- sim_data$simmed_data %>% select(S_obs, Y) 
med_risk_cross_sec_results <- Thres_CrossSect(dataset = vac_data_analysis_MR, p=c(0.009, 0.01, 0.03, 0.05, 0.07, 0.09), weighted=TRUE, bts=2000)
med_risk_cross_sec_results

# Now plot
plotCI(x=med_risk_cross_sec_results$Risk_Level, y=med_risk_cross_sec_results$Est, 
       ui=med_risk_cross_sec_results$CI_UL, li=med_risk_cross_sec_results$CI_LL, xlab="Risk Level",
       ylab="Threshold Estimate",col="red",scol="black", main="Threshold of risk inference on simulated data: disease risk=0.1\nTime to event scenario")
points(x=as.numeric(names(sim_data$true_thresholds)), y=sim_data$true_thresholds, col="blue", sub="Cross sectional scenario")
```

## Zero risk level
For the cross sectional data scenario, suppose $\Pr(Y=1|S)=$exbit$(\beta_0+\beta_1S)$ for $S<v_0$ and $\Pr(Y=1|S)=0$ otherwise (as described in Section 2 of the paper).  In the following simulated corss-sectional examples, everything else with the previous cross-sectional simulations is unchanged.  The R function sim_data_logit found in the GitHub repository simulates data under this model using the argument **theta** (denotes the value of $v_0$ to simulate data under).  The nonparametric estimator $\tilde{v}_0$ is computed, with the corresponding bootstrap confidence intervals as detailed in Section 2 of the paper.

The two simulated datasets used in the following examples, one simulated under a 0.01 disease risk and the other under 0.10 disease risk with $v_0$=2 for both scenarios, using the function sim_data_logit provided in the GitHub repository.

```{r cross_sec_zero_logit}
source("../sim_data_logit.R")
source("../Thres_CrossSect.R")

## Low risk example
sim_data <- Sim_data_logit(Risk = 0.01, n=12500, theta=2, p0=1, p1=1, p=0, seed_val = 012)

vac_data_analysis_LR <- sim_data$simmed_data %>% select(S_obs, Y) 
low_risk_cross_sec_results <- Thres_CrossSect(dataset = vac_data_analysis_LR, p=0, weighted=FALSE, bts=2000)
low_risk_cross_sec_results

## Medium risk example
sim_data <- Sim_data_logit(Risk = 0.1, n=12500, theta=2, p0=0.2, p1=1, p=0,seed_val = 012)

vac_data_analysis_MR <- sim_data$simmed_data %>% select(S_obs, Y) 
med_risk_cross_sec_results <- Thres_CrossSect(dataset = vac_data_analysis_MR, p=0, weighted=FALSE, bts=2000)
med_risk_cross_sec_results
```

# Time to Event Data
Now, let's consider the scenario where inference on risk thresholds is done with time to event data (as described in Section 4 of the paper).  Let time $t_0$ denote end of follow-up ($t_0>0$), $T$ denote survival time, $S$ denote immune response, $C$ denote censoring status ($C=1$ denoting right censoring at or before time $t_0$, and $C=0$ denoting observed disease before time $t_0$).  Suppose $S$ is observed for proportion $p_0$ of subjects with $C=1$ and proportion $p_1$ of subjects with $C=0$.  In the following simulated examples, sample size $n=12,500$ where $S$ is generated from a gamma distribution with mean and variance 4 and $T|S$ is generated by log$(T)=\beta_0+\beta_1S+\epsilon$ where $\epsilon$ is generated from a logistic distribution.  The slope $\beta_1$ sas set to 2 and $\beta_0$ was determined based on the chosen disease risk $\Pr(T<t_0)$; in the following simulations risks of 0.01 and 0.1 were used with $p_0=0.2$ and $p_1=1$.  The R function sim_data_logistic found in the GitHub repository simulates data under this model (based on choosen values of $n, \beta_1, t_0$, $p_0$, $p_1$, and disease risk).  

For disease risk 0.01, point estimates and 95$\%$ confidence intervals are computed of thresholds for risk levels 0.001, 0.003, 0.005, 0.007, and 0.009 and for disease risk 0.1, point estimates and 95$\%$ confidence intervals are computed of thresholds for risk levels 0.009, 0.01, 0.03, 0.05, 0.07, and 0.09.  The confidence intervals are computed using 2000 bootstrap resamples.  These calculations were done using the R function Thres_Surv, whose source code can be found in the GitHub repository, which implements the methods detailed for Section 4 for inference with time to event data.  These estimates and confidence intervals were computed using $\hat{v}_c^{w}$ (referred to as the "naive" method in the R function Thres_Surv) and $\hat{v}_c^{wk}$ ("adjusted" method), as defined in Section 4 of the paper.  For each threshold, the point estimate and confidence interval using $\hat{v}_c^{wk}$ are plotted with the corresponding risk level on x axis; the point estimate is colored red and the true threshold is colored blue.

The true thresholds under the model from which the data were simulated (models detailed previously) are also provided (function sim_data_logistic calculates these for vector of choosen risk levels).

```{r time_to_event_logistic}
source("../sim_data_logistic.R")
source("../Thres_Surv.R")

## Low risk example
sim_data <- Sim_data_logistic(Risk = 0.01, n=12500, p=c(0.001,0.003,0.005,0.007,0.009), p0=0.2, p1=1, seed_val = 012)

vac_data_analysis_LR <- sim_data$simmed_data %>% select(S_obs, TObs, Cstatus) 
low_risk_tte_results <- Thres_Surv(dataset = vac_data_analysis_LR, p=c(0.001,0.003,0.005,0.007,0.009), weighted=TRUE, bts=2000)
low_risk_tte_results
adjusted_results_low <- low_risk_tte_results %>% filter(Method=="Adjusted")

# Now plot
plotCI(x=adjusted_results_low$Risk_Level, y=adjusted_results_low$Est, 
       ui=adjusted_results_low$CI_UL,li=adjusted_results_low$CI_LL, xlab="Risk Level", ylab="Threshold Estimate",
       col="red",scol="black", main="Threshold of risk inference on simulated data: disease risk=0.01\nTime to event scenario")
points(x=as.numeric(names(sim_data$true_thresholds)), y=sim_data$true_thresholds, col="blue")

## Medium risk example
sim_data <- Sim_data_logistic(Risk = 0.1, n=12500, p=c(0.009, 0.01, 0.03, 0.05, 0.07, 0.09), p0=0.2, p1=1, seed_val = 012)

vac_data_analysis_MR <- sim_data$simmed_data %>% select(S_obs, TObs, Cstatus) 
med_risk_tte_results <- Thres_Surv(dataset = vac_data_analysis_MR, p=c(0.009, 0.01, 0.03, 0.05, 0.07, 0.09), 
                                   weighted=TRUE, bts=2000)
med_risk_tte_results
adjusted_results_med <- med_risk_tte_results %>% filter(Method=="Adjusted")

# Now plot
plotCI(x=adjusted_results_med$Risk_Level, y=adjusted_results_med$Est, 
       ui=adjusted_results_med$CI_UL, li=adjusted_results_med$CI_LL, xlab="Risk Level", ylab="Threshold Estimate",col="red",scol="black", main="Threshold of risk inference on simulated data: disease risk=0.1\nTime to event scenario")
points(x=as.numeric(names(sim_data$true_thresholds)), y=sim_data$true_thresholds, col="blue", sub="Time to event scenario")

```
